### 2022-06-12 paper.tex
marked notes in orange

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ intead. But we don't do that, right?

>>>>

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is then to assigning cell types or states to cells, and to this end, one needs to establish which cells are similar to each other, i.e., find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation y are suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

\todo{NOTE: Here, we could already do shrinkage and use $\frac{1}{|C_i|+1}\sum_i x_{ij}$ intead. But we don't do that, right?}
